\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\newtheorem{ex}{Example}
\newtheorem{axm}{Axiom}

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{prop}{Proposition}

\title{The Method of Bisection}
\date{\today}
\author{Jonathan Lin}

\begin{document}
\maketitle
\begin{abstract}
I summarize a method called the ``Bisection Argument'', which can be used to prove several important results in real analysis.
\end{abstract}

\section{The Nested Interval Theorem}
All these results rely on the following theorem:
\begin{thm}
	Suppose that $I = [a, b]$ is an interval. If we have a series of intervals $I_n$ such that $I_0 = [a, b]$ and $I_n = [a_n, b_n]$ where we have the chain
	\[I_0 \supset I_1 \supset I_2 \supset \cdots.\]

	Suppose also that 
	\[\lim_{n \to \infty}a_n - b_n = 0.\]
	Then there is a unique point $c$ contained in the $I_n$.
\end{thm}

If we let go of the requirement that $\lim_{n \to \infty} a_n - b_n = 0$, then the point $c$ still exists, but it's not necessarily unique (in fact, there will be uncountably many $c$.

\begin{proof}
The sequence $\{a_n\}$ is a monotonically increasing sequence which is bounded above by any $b_n$. Similarly, the sequence $\{b_n\}$ is a monotonically decreasing sequence which is bounded below by any $a_n$. It follows both limits exist, and we have
\[\lim_{n \to \infty} a_n = \sup(A)\] and \[\lim_{n \to \infty}b_n = \inf(B).\] Since both limits are equal (by hypothesis) we have that their limit is the unique point contained in all the $I_n$ (follows from consequences of the least upper bound).
\end{proof}

\section{Bisection of an Interval}

Here is the approach: Suppose we want to establish the existence of a point in a closed interval $[a, b]$ with property $P$. A priori we know this proposition to be true. We would like to ``seek out'' the point in question. What we might do is divide our interval in half. Then we might be able to deduce that in all cases the property $P$ holds for either $\left[a, \frac{a + b}{2}\right]$ or $\left[\frac{a + b}{2}, b\right]$. If we iteratively do this, we end up with intervals which are contained in each other with length $2^{-n}(b-a)$. After this we will be able to apply the nested interval theorem to this sequence of intervals. We summarize these above remarks by defining a \textbf{bisection} below.

\begin{defn}
	Let $I = [a, b]$ and let $P$ be some property which holds for $I$, and also has the property that it holds for $\left[a, \frac{a + b}{2}\right]$ or $\left[\frac{a + b}{2}, b\right]$. We define $I_0 = 0$ and define a sequence $I_n$ inductively as follows: given $I_n = [a_n, b_n]$ for which the property $P$ holds, then $P$ holds for either $\left[a, \frac{a_n + b_n}{2}\right]$ or $\left[\frac{a_n + b_n}{2}, b\right]$ in which case we define $I_{n + 1}$ to be the interval for which the property $P$ holds. We call this sequence a \textbf{bisection} of $I$.
\end{defn}

Note that any bisection of $I$ has the following properties: We have the chain
\[I = I_0 \supset I_1 \supset \cdots \supset I_n \supset \cdots\]
and if $I_n = [a_n, b_n]$ we have that $b_n - a_n = (b-a)2^{-n}$, so we have that 
\[\lim_{n \to \infty}b_n - a_n = 0.\] It follows that the nested interval theorem can be applied to the bisection of $I$ to get a point $c$ contained in all the $I_n$.

\section{The Intermediate Value Theorem}

\begin{thm}
Suppose $f$ is continuous on an interval $[a, b]$. Also, suppose that $f(a) \leq 0 \leq f(b)$. It follows that there exists a point $c$ in $[a, b]$ such that $f(c) = 0$.
\end{thm}

\begin{proof}
Let $P$ be the property on an interval $[a, b]$ such that $f$ is continuous on that interval and $f(a) \leq 0 \leq f(b)$. The point $(a + b)/2$ is either non-negative or non-positive. This means that $P$ holds on one of the half-intervals of $[a,b]$. Inductively, this holds as well. Hence we can construct a bisection $\{I_n\}$ where $P$ holds all the $I_n$. By the nested interval theorem there is a unique point $c$ contained in the $I_n$.

We claim that $f(c) = 0$. We can see this as follows: Since $c$ is contained in the $I_n$ we have that $f(a_n) \leq 0$ for all $a_n$ and $f(b_n) \geq 0$ for all $b_n$. Since $f$ is continuous, we have that $f(a_n) \to f(c)$ and $f(b_n) \to f(c)$. Combining these facts, we have that $f(c) \leq 0$ and $f(c) \geq 0$, implying that $f(c) = 0$.
\end{proof}

\section{The Extreme Value Theorem}

\begin{thm}
If $f$ is continuous on $[a, b]$ then $f$ is bounded above on $[a, b]$. Moreover, $f$ attains its supremum, so that $f$ has a maximum value.
\end{thm}

\begin{proof}
Suppose that $f$ is unbounded on $[a, b]$. Then $f$ is unbounded on at least one of its half-intervals. We define the bisection $\{I_n\}$ with the property that $I_n$ is unbounded. By the nexted interval theorem, we have that 
\[\bigcap_{i = 1}^{\infty}I_n = \{c\}\] for $c \in [a, b]$. But since $f$ is continuous at $c$, there exists a $\delta > 0$ such that $|x - c| < \delta \implies |f(x) - f(c)| < 1$. So $f$ is bounded on $(c - \delta, c + \delta)$. HOwever, there exists $N \in \mathbb{N}$ such that $I_N \subset (c - \delta, c + \delta)$, which is a contradiction because $I_N$ is an unbounded set.

Clearly then we have that $S = \{f(x) \mid x \in [a,b]\}$ is bounded and non-empty. Hence $\alpha = \sup(S)$ exists. Clearly either one of the half-intervals of $[a, b]$ must have supremum $\alpha$ as well. We define the bisection $\{I_n\}$ with the property that $sup(f(I_n)) = \alpha$. By the nested interval theorem, we have that there is a unique common point $\ell$ in the $I_n$. We claim that $f(\ell) = \alpha$. This is because if $f(\ell) \neq \alpha$, then if $\varepsilon = \frac{1}{2}(\alpha - f(\ell))$ there would exist a $\delta > 0$ such that $|x - \ell| < \delta \implies |f(x) - f(\ell)| < \varepsilon$. However, there is an interval $I_n$ inside $(x - \delta, x + \delta)$, which is a contradiction since that would mean there is a value less than $\alpha$ which is an upper bound for $I_n$.
\end{proof}

\section{An application to Uncountability}

We will show that the interval $[0, 1]$ is uncountable. This argument is identical to the Cantor diagonalization argument, but the argument is made using the nested interval theorem and repeated bisections. First, note that our bisection will be defined as a trisection of the first and last third intervals. This is so we can use the nested interval theorem and there are no intersections.

\begin{thm}
The interval $[0, 1]$ is uncountable. That is, there is no bijection from $[0, 1]$ to the natural numbers $\mathbb{N}$.
\end{thm}

\begin{proof}
Suppose that $[0, 1]$ is countable. That is, there was some list $\{x_n\}$ of all natural numbers under some ordering. Define the bisection $I_n$ to be the lower third of $I_n$ if $x_n$ is not in that interval and the upper third of $I_n$ otherwise.

By the nested interval theorem, the $I_n$ contain a unique point $x$. However, x is contained in the $I_n$, but none of the $x_i$ are contained in all of the $I_n$. In particular, $x_j \not\in I_j$ for all $j \in \mathbb{N}$. This is a contradiction, so it follows that $[0, 1]$ is uncountable.
\end{proof}
\end{document}
